Variance reduction for Markov chains via martingale representations

### 2020-05-09 aap/Revision/vr-mcmc-revision-final.tex
update

If the class of functions $\Psi$ is a linear one of dimension $\mathtt{D}_\Psi,$ then the cost of computing the coefficients $\widehat A_{l,k}$ for all $l=1,\ldots,n,$ and $k=1,\ldots,K,$ is of order  $ \mathtt{D}_\Psi K T^2 n$. Given $(\widehat A_{l,k})$ the cost of computing  $\widehat \pi_{K,n}^N$ is proportional to $\mathtt{D}_\Psi K n.$ At the same time the variance of the standard estimate $ \pi_{n}^N(f)$ is of order $1/(n\gamma)$ and this bound can not be improved, see Lemma~\ref{lem:variance} and Remark~\ref{rem:var-low}. Thus, the ratio of the corresponding cost-to-variance characteristics  is of order

@@

Thus, for any fixed $K\geq 1,$  our variance reduction method is advantageous if $\gamma\ll\min\{1/(\mathtt{D}_\Psi T^2),1/\sqrt{n}\}.$  Note that in order to achieve convergence of the corresponding MSE to zero, we need to let $\gamma\to 0$ as the invariant measure of the ULA with a constant time step $\gamma$ is not equal to $\pi.$

@@

\item[(H2)][\textbf{Convexity}] The potential $U$ is of the form $U(x)=U_0(x)+\Delta(x),$ where $\Delta$ has  compact support and $U_0\in C^2(\mathbb{R}^d)$ with

@@

@@

\gamma\sum_{l=j+1}^{p}\prod_{k=j+1}^{l-1}\left(I-\gamma\mathsf{D}^{2}U_{0}(X_{k-1})\right)\mathsf{D}^{2}\Delta(X_{l-1})\prod_{k=l+1}^{p}\left(I-\gamma\mathsf{D}^{2}U_0(X_{k-1})\right),

@@

&\leq &  \left(1+\frac{\gamma (p-j) D}{1-\gamma\mathfrak{M}}\right) (1-\gamma \mathfrak{m})^{p-j}.

@@

@@

A^2_{q,\mathbf{k}}(x)\leq \gamma^{|\mathbf{k}|} \PVar \left( \sum_{p=1}^{q}\partial_{Z_{1}}^{\mathbf{K}_I}f\left(X_{p}^x\right)

@@

\PVar \left( \sum_{p=1}^{q}\partial_{Z_{1}}^{\mathbf{K}_I}f\left(X_{p}^x\right)
\right),

@@

\PVar \left(\sum_{p=1}^{q}\partial_{Z_{1}}^{\mathbf{K}_I}f\left(X^x_{p}\right)

@@

for some positive constants $\varkappa, C_K$ not depending on $n$ and  $\gamma.$

>>>>

If the class of functions $\Psi$ is a linear one of dimension $\mathtt{D}_\Psi,$ then the cost of computing the coefficients $\widehat A_{l,k}$ for all $l=1,\ldots,n,$ and $k=1,\ldots,K,$ is of order  $ \mathtt{D}_\Psi K T^2 n$. Given $(\widehat A_{l,k})$ the cost of evaluating  $\widehat \pi_{K,n}^N$ is proportional to $\mathtt{D}_\Psi K n.$ At the same time the variance of the standard estimate $ \pi_{n}^N(f)$ is of order $1/(n\gamma)$ and this bound can not be improved, see Lemma~\ref{lem:variance} and Remark~\ref{rem:var-low}. Thus, the ratio of the corresponding cost-to-variance characteristics  is of order

@@

Thus, for any fixed $K\geq 1,$  our variance reduction method is advantageous if $\gamma\ll\min\{1/(\mathtt{D}_\Psi T^2),1/\sqrt{n}\}.$  Note that in order to achieve convergence of the corresponding MSE to zero, we need anyway that $\gamma\to 0$ as the invariant measure of the ULA with a constant time step $\gamma$ is not equal to $\pi.$

@@

\item[(H2)][\textbf{Convexity}] The potential $U$ has the form $U(x)=U_0(x)+\Delta(x)$ where $\Delta$ has  a compact support and $U_0\in C^2(\mathbb{R}^d)$ with

@@

 \label{eq:cor-bound}

@@

\gamma\sum_{l=j+1}^{p}\prod_{k=j+1}^{l-1}\left(I-\gamma\mathsf{D}^{2}U_{0}(X_{k-1})\right)\mathsf{D}^{2}\Delta(X_{l-1})\prod_{k=l+1}^{p}\left(I-\gamma\mathsf{D}^{2}U_0(X_{k-1})\right)

@@

&\leq &  \left(1+\frac{\gamma (p-j) D}{1-\gamma\mathfrak{M}}\right) (1-\gamma \mathfrak{m})^{p-j}

@@

and  \eqref{eq:cor-bound} follows.

@@

A^2_{q,\mathbf{k}}(x)\leq \gamma^{|\mathbf{k}|} \PVar \left( \sum_{p=0}^{q}\partial_{Z_{1}}^{\mathbf{k}}f\left(X_{p}^x\right)

@@

\PVar \left( \sum_{p=0}^{q}\partial_{Z_{1}}^{\mathbf{K}_I}f\left(X_{p}^x\right)

@@

\PVar \left(\sum_{p=1}^{q}\partial_{Z_{1}}^{\mathbf{k}}f\left(X^x_{p}\right)

@@

for any $\| \mathbf{k} \| \leq K$ and some positive constants $\varkappa, C_K$ not depending on $n$ and  $\gamma.$
